---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# streamflow

The streamflow package let's users classify streams as being wet or dry using time-series data generated by sensors. The package should be considered an _alpha_ release but we would love to hear any feedback, feature requests, or bug reports that you might encounter. 

## Installation

The streamflow package is not yet on CRAN but you can download and install the current working version from github using the following:

```r
# Install the remotes package if you haven't already.
# install.packages(remotes)
remotes::install_github("jimmylovestea/streamflow")
```

After installing, load the package into your R environment using

```{r}
library(streamflow)
```


## Example usage

The package includes a dataset from the Willow Whitehorse creek area in Oregon. The data includes resistor and water temperature sensor readings for 10 different sites recorded in 2015 between the Spring and late Summer.

For this data, the first 6 rows look as follows:

```{r head-data, echo = FALSE}
knitr::kable(head(wwc))
```

A few things to notice here are that the `date_time` variable is a `POSIX` class variable, meaning it has information about both the date and time of the sensor reading. Also of imporance is to note that there are two types of sensors in the `sensor_type`. The names of these sensors are

```{r sensor-types}
unique(wwc$sensor_type)
```

Finally, note that the initial and end state of the stream is noted in the data but this information is not required for the functions to work.

Once your raw data has a similar structure as the example data, you'll need to further format your data in order for the classification methods to work. To do so, you can run:

```{r}
formatted_wwc <- format_data(data = wwc,
                             date_var = "date_time",
                             date_start = "2015-04-01",
                             date_end = "2015-08-30",
                             site_id_var = "site_id",
                             sensor_type_var = "sensor_type",
                             resistor_level = "Resistor",
                             water_temp_level = "Water temp",
                             sensor_reading_var = "sensor_reading")
```

The function arguments `date_var`, `site_id_var`, `sensor_type_var`, and `sensor_reading_var` should each be the name of the variables in the data which contain their respective information. You can subset your data to include specific time-periods by specifying `date_start` and `date_end`. You'll also need to include the names of the sensors from `sensor_type_var` in the `resistor_level` and `water_temp_level` arguments.

Once your data is formatted, it will be formatted so that each row represents one days worth of sensor readings for each sensor. It will also include the mean resistor reading for the day along with the hourly temperature readings from the water temperature sensor as columns.

Right now, the package ships with one classification method (though more are in the works). This classification method requires _centers_ for the wet and dry streams to be computed. To compute these centers, run:

```{r compute-centers}
centers <- compute_centers(data = formatted_wwc, daily_mean_resistor = "daily_mean_resistor")
centers
```

The _centers_ are used in the classification method and both the wet and dry centers cover the 24 hours of a day.

Having computed the centers, we can now classify the data using:

```{r classify-data-numeric}
classifications <- classify_streams(data = formatted_wwc, centers = centers)
head(classifications)
```

The returned values by default are numeric where values greater than one indicate the stream was "Dry" and values less than one indicate the stream was "Wet". Alternatively, you can have the "Wet" and "Dry" labels returned directly by specifying `labels = TRUE`:

```{r classify-data-labels}
classifications <- classify_streams(data = formatted_wwc, centers = centers, labels = TRUE)
head(classifications)
```

## Summarizing the classifications

For the Willow Whitehorse Creek data, we can verify the performance of the classification method by comparing the output of `classify_streams` to the resistor sensor readings:

```{r}
# Compute the predicted status of the stream using the classification method
predicted_status <- classify_streams(data = formatted_wwc, centers = centers, labels = TRUE)

# Compute the acutal status of the stream using the resistor readings
actual_status <- ifelse(formatted_wwc$daily_mean_resistor >= 0, "Wet", "Dry")

# Compare the predicted and actual status readings
table(actual_status, predicted_status)
```

## Future development plans

In the future, we plan to include in this package:

- Additional classification methods and additional methods to compute the _centers_ of the time series.
- Plotting and summary methods to more easily summarize the results for users.

If there's methods you would like to see implemented, drop us an email or file an issue on the Github repo.
